Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-05-10 doc/manuscript/translocation_pnas_SI.tex

\title{Reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease}
\author{Roland A. Knapp, Mark Q. Wilber, Allison Q. Byrne, Maxwell B. Joseph, Thomas C. Smith, Andrew P. Rothstein, Robert L. Grasso, Erica Bree Rosenblum}
\correspondingauthor{Corresponding Author name. Roland A. Knapp \\ E-mail: roland.knapp@ucsb.edu}

@@

@@

\caption{\label{fig-selectionresults}Results from the exome capture
selection study. (A) PCA calculated from binary SNPs showing genomic
relationship of samples. Numeric labels and colors match those in
Fig. 1 A. (B) qqplot showing observed and
expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
Dashed line shows p-value that identifies outliers. (C) Manhattan plot
showing p-values for each SNP as calculated by GEMMA. SNPs are sorted by
genomic position and contigs are sorted by size. Red line shows p-value
that identifies outliers. Outlier SNPs above this threshold are
highlighted and labeled. Bold labels indicate at least one
non-synonymous SNP is present in that gene.\\
}

@@

\caption{\label{fig-bdload-beforeafter}For frogs translocated to each of
the 12 recipient sites, Bd loads for the period immediately prior to
translocation versus during the 1-year period after translocation. Box
plot details are as in Fig. 4.}

@@

\hypertarget{tbl-param_values}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4545}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1333}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4121}}@{}}
\caption{\label{tbl-param_values}Description and values of parameters
used in the model. All survival probabilities are in the presence of the
fungal pathogen Bd.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Parameter}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Value}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Source}
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Parameter}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Value}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Source}
\end{minipage} \\
\midrule()
\endhead
\(\sigma_{L_1}\), Yearly survival probability of year 1 tadpoles & 0.7 &
Estimated from field data, observations, natural history knowledge \\
\(\sigma_{L_2}\), Yearly survival probability of year 2 tadpoles & 0.7 &
Estimated from field data, observations, natural history knowledge \\
\(\sigma_{L_3}\), Yearly survival probability of year 3 tadpoles & 0.7 &
Estimated from field data, observations, natural history knowledge \\
\(\sigma_{J_1}\), Yearly survival probability of year 1 juveniles & 0.25
& Estimated from field data, observations, natural history knowledge \\
\(\sigma_{J_2}\), Yearly survival probability of year 2 juveniles & 0.5
& Estimated from field data, observations, natural history knowledge \\
\(\sigma_{A_R}\), Yearly survival probability of naturally recruited
adults & Varies by population & Estimated from mark-recapture studies \\
\(\sigma_{A_T}\), Yearly survival probability of translocated adults &
Varies by population & Estimated from mark-recapture studies \\
\(p_{L_1}\), Probability of a year 1 tadpoles remaining as a tadpoles &
1 & Estimated from field data, observations, natural history
knowledge \\
\(p_{L_2}\), Probability of a year 2 tadpoles remaining as a tadpoles &
0.25 & Estimated from field data, observations, natural history
knowledge \\
\(p_{J_1}\), Probability of a year 1 juvenile remaining as a juvenile &
0.25 & Estimated from field data, observations, natural history
knowledge \\
\(p_F\), Probability of a adult female reproducing in a year & 0.5 &
Could be as high at 1, based on field observations \\
\(F\), Number of surviving eggs produced by an adult female & 100 & From
observations of captive frogs \\
\(\omega\), Probability of juvenile successfully recruiting to an adult
& Varies yearly & Explored different values of this parameter \\
\bottomrule()
\end{longtable}

>>>>

\title{Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease}
\author{Roland A. Knapp, Mark Q. Wilber, Allison Q. Byrne, Maxwell B. Joseph, Thomas C. Smith, Andrew P. Rothstein, Robert L. Grasso, Erica Bree Rosenblum}
\correspondingauthor{Corresponding Author name. Roland A. Knapp \\ E-mail: roland.knapp@ucsb.edu}

@@

\hypertarget{frog-population-recovery-2}{%
\section{Frog population
recovery}\label{frog-population-recovery-2}}

\hypertarget{methods-1}{%
\subsection{Methods}\label{methods-1}}

Swab extracts were analyzed singly instead of in triplicate
\citep{kriger2006}. For analysis of swabs collected during 2005--2014,
we used standards developed from known concentrations of zoospores
\citep{boyle2004}, and after 2014, we used standards based on single
ITS1 PCR amplicons \citep{longo2013}. Based on paired comparisons
between samples analyzed using both types of standards, Bd in the study
area has an average of 60 ITS1 copies per zoospore. To express all qPCR
results as the number of ITS1 copies, starting quantities obtained using
the zoospore standard (measured as ``zoospore equivalents'') were
multiplied by 60. In addition, all qPCR quantities (regardless of
standard) were multiplied by 80 to account for the fact that DNA
extracts from swabs were diluted 80-fold during extraction and PCR
\citep{vredenburg2010}.

\hypertarget{population-viability-modeling-1}{%
\section{Population viability
modeling}\label{population-viability-modeling-1}}

\hypertarget{methods-2}{%
\subsection{Methods}\label{methods-2}}

\hypertarget{incorporating-yearly-variability-in-vital-rates}{%
\subparagraph{Incorporating yearly variability in vital
rates}\label{incorporating-yearly-variability-in-vital-rates}}

We extracted yearly survival probabilities for translocated adults
\(\sigma_{A_T}\) and naturally recruited adults \(\sigma_{A_R}\) from
the CMR model used to estimate population-level parameters of
translocated populations (see ``Estimation of adult survival and
population size'' above). Although we observed yearly variability in
adult survival within a population, the magnitude of this variability
was small compared to among population variability
(Fig. 3). Thus, we did not include
yearly within-population variability in adult survival in this analysis.
However, within a population there was substantial yearly variability in
the successful recruitment of adults, greater than what we would expect
from Poisson variability around a mean value. Therefore, we allowed for
yearly variability in the probability of successful recruitment
\(\omega\) (additional details provided in ``Estimating model
parameters'' below).

\hypertarget{estimating-model-parameters}{%
\subparagraph{Estimating model
parameters}\label{estimating-model-parameters}}

The baseline parameter values for the model and how they were estimated
are given in Table~\ref{tbl-param_values} : SI. Parameters
\(\sigma_{A_T}\) and \(\sigma_{A_R}\) were extracted directly from our
CMR model described previously in this study. For populations where we
had a sufficient number of PIT-tagged, naturally-recruited adults, we
observed that \(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be notably
different, with \(\sigma_{A_R} > \sigma_{A_T}\)
(Figure~\ref{fig-compare_surv_probs} : SI). For populations lacking
sufficient numbers of naturally-recruited adults, we were unable to
directly estimate \(\sigma_{A_R}\), and instead set
\(\sigma_{A_R} = \sigma_{A_T}\).

\hypertarget{model-analysis-and-simulation}{%
\subparagraph{Model analysis and
simulation}\label{model-analysis-and-simulation}}

We performed four analyses on our model. First, we considered a
deterministic version of our model with no yearly heterogeneity in
probability of successful recruitment \(\omega\), and calculated the
predicted long-run growth rate \(\lambda\) of a population for different
values of \(\sigma_{A_R}\) and \(\omega\). We then fixed
\(\omega = 0.3\) and calculated the predicted growth rate of our 12
populations.

Second, we performed a local elasticity analysis on \(\lambda\) with
respect to parameters \(\omega\), \(\sigma_{A_R}\), and \(F\) to
determine how small changes in these parameters could influence the
long-run deterministic growth rate of populations.

Third, we defined a version of the model with demographic and
environmental stochasticity, where environmental stochasticity was
represented by among-year variability in \(\omega\). We used this model
to simulate a one-time introduction of 40 translocated adult frogs. We
ran this simulation 1000 times for each population and computed the
probability of a population becoming extinct after 50 years given the
observed parameter values and environmental stochasticity in \(\omega\).
We varied the mean recruitment probability \(\omega\) from 0.15 and 0.4
and drew values of \(\omega\) each year from a beta distribution with a
dispersion parameter of \(\phi = 2\) (when \(\omega = 0.5\) and
\(\phi = 2\) the beta distribution is uniform between 0 and 1).

Finally, we assessed whether our stochastic model could reproduce
observed trajectories of population recovery. We focused on population
70550 because this was our longest CMR time series for a translocated
population and because this population shows evidence of substantial
post-translocation increases in adult abundance associated with
population establishment and recovery. We simulated our model for 16
years, repeating the simulation 50,000 times. For each run and each
year, we drew \(\omega\) from a uniform distribution between 0 and 1 (or
equivalently a beta distribution with mean 0.5 and \(\phi = 2\)). Using
Approximate Bayesian Computing and rejection sampling
\citep{kosmala2016}, we identified the top 2\% of trajectories (i.e.,
1000 trajectories) that minimized the sum of squared errors between the
observed and predicted data. The \(\omega\) values associated with these
``best'' trajectories represented an approximate posterior distribution
\citep{beaumont2010}. Using these best fit trajectories, we assessed
whether our model could qualitatively describe the patterns of recovery
in the observed data for population 70550.

@@

\caption{\label{fig-selectionresults}Results from the exome capture
selection study. (A) PCA calculated from binary SNPs showing genomic
relationship of samples. Numeric labels and colors match those in
Fig. 1 A. (B) qqplot showing observed and
expected p-values for 148,307 SNPs and indels as calculated in GEMMA.
Dashed line shows p-value that identifies outliers. (C) Manhattan plot
showing p-values for each SNP as calculated by GEMMA. SNPs are sorted by
genomic position and contigs are sorted by size. Red line shows p-value
that identifies outliers. Outlier SNPs above this threshold are
highlighted and labeled. Bold labels indicate at least one
non-synonymous SNP is present in that gene.}

@@

\caption{\label{fig-bdload-beforeafter}For frogs translocated to each of
the 12 recipient sites, Bd loads for the period immediately prior to
translocation versus during the 1-year period after translocation. Bd
loads are expressed as the number of ITS1 copies per skin swab, as
estimated by qPCR of the Bd ITS1 region. Loads indicative of severe
disease are \textgreater{} 5.8 ITS copies (on a log\textsubscript{10}
scale; see SI text for details). Box plot details are as in
Fig. 4.}

@@

\hypertarget{tbl-param_values}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3750}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2639}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}@{}}
\caption{\label{tbl-param_values}Description and values of parameters
used in the model. All survival probabilities are in the presence of the
fungal pathogen Bd.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Parameter}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Value}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Source}
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Parameter}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Value}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Source}
\end{minipage} \\
\midrule()
\endhead
\(\sigma_{L_1}\), Yearly survival probability of year 1 tadpoles & 0.7 &
Estimated from field data, observations, natural history knowledge \\
\(\sigma_{L_2}\), Yearly survival probability of year 2 tadpoles & 0.7 &
Estimated from field data, observations, natural history knowledge \\
\(\sigma_{L_3}\), Yearly survival probability of year 3 tadpoles & 0.7 &
Estimated from field data, observations, natural history knowledge \\
\(\sigma_{J_1}\), Yearly survival probability of year 1 juveniles & 0.25
& Estimated from field data, observations, natural history knowledge \\
\(\sigma_{J_2}\), Yearly survival probability of year 2 juveniles & 0.5
& Estimated from field data, observations, natural history knowledge \\
\(\sigma_{A_R}\), Yearly survival probability of naturally recruited
adults & Varies by population & Estimated from mark-recapture studies \\
\(\sigma_{A_T}\), Yearly survival probability of translocated adults &
Varies by population & Estimated from mark-recapture studies \\
\(p_{L_1}\), Probability of a year 1 tadpoles remaining as a tadpoles &
1 & Estimated from field data, observations, natural history
knowledge \\
\(p_{L_2}\), Probability of a year 2 tadpoles remaining as a tadpoles &
0.25 & Estimated from field data, observations, natural history
knowledge \\
\(p_{J_1}\), Probability of a year 1 juvenile remaining as a juvenile &
0.25 & Estimated from field data, observations, natural history
knowledge \\
\(p_F\), Probability of a adult female reproducing in a year & 0.5 &
Could be as high at 1, based on field observations \\
\(F\), Number of surviving eggs produced by an adult female & 100 & From
observations of captive frogs \\
\(\omega\), Probability of juvenile successfully recruiting to an adult
& Varies yearly & Explored different values of this parameter \\
\bottomrule()
\end{longtable}
